---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setdefaults,results='hide'}
library(knitr)
opts_chunk$set(echo = TRUE,cache=TRUE)
```
```{r loadpackages,message=FALSE}
library(dplyr)
library(lattice)
```
## Loading and preprocessing the data
1. loading the data from _"activity.zip"_ file :
```{r loading}
activity <- read.csv(unz("activity.zip","activity.csv"))
```

2. processing the data as required by us :    
  `date field is still as character lets make it Date class.`   
```{r processing}
activity$date <- as.Date(activity$date)
```
## What is mean total number of steps taken per day?
1. Histogram of total number of steps taken each day :
```{r histogram,fig.align='center',message=F}
by_date <- activity %>% 
           group_by(date) %>%
           summarise(tot_st = sum(steps,na.rm=T))
hist(by_date$tot_st,xlab = "Number of steps",main ="Total steps taken each Day" )
```

2. Calculating mean and median total number of steps taken per day :
```{r mean&median}
mean_bydt <- mean(by_date$tot_st)
med_bydt <- median(by_date$tot_st)
summary(by_date$tot_st)
```
The mean and median of total number of steps taken each day are `r round(mean_bydt,0)` and `r med_bydt`
respectively.  

## What is the average daily activity pattern?

1.Time series plot of the 5-minute interval and the average number of steps taken, averaged across all days.
```{r by_interval,message=FALSE,fig.align='center'}
by_inter <- activity %>% 
           group_by(interval) %>%
           summarise(tot_st = mean(steps,na.rm=T))
plot(by_inter$interval,by_inter$tot_st,type="l",xlab = "interval",ylab = "steps")
title(main = "Average number of steps in interval")
```

2.The time interval with maximum average steps :
```{r max_steps}
maxI <- by_inter$interval[which.max(by_inter$tot_st)]
print(maxI)
```
So, maximum steps are taken at around `r round(maxI/60,0)`:00 hours daily.  

## Imputing missing values  
There are a number of days/intervals where there are missing values (coded as `NA`). The presence of missing days may introduce bias into some calculations or summaries of the data.  

```{r na_no}
nano <- sum(is.na(activity$steps))
print(nano)
```
1. The total number of missing values in the dataset is `r nano`, which is `r round(nano/nrow(activity),2)*100`% approximately.

2. Lets impute these `NA` values with average steps in each interval across all days.This seems like a valid replacement considering daily schedule of a person generally remains same.  
```{r impute}
imputed <- activity %>%
           group_by(interval) %>%
           mutate(avg_st = mean(steps,na.rm=T),steps = ifelse(is.na(steps),avg_st,steps)) %>%
           select(steps,date,interval)
```


3. Now we have a new dataset with imputed values for `NAs`.  

4. Lets see previous results for this dataset now.

* Histogram of total number of steps taken each day :
```{r hist_new,fig.align='center',message=F}
by_date_imp <- imputed %>% 
               group_by(date) %>%
               summarise(tot_st = sum(steps,na.rm=T))
hist(by_date_imp$tot_st,xlab = "Number of steps",main ="Total steps taken each Day" )
```

* Means and medians of total number of steps each day:
```{r summ_imp}
mean_imp <- mean(by_date_imp$tot_st)
med_imp <- median(by_date_imp$tot_st)
summary(by_date_imp$tot_st)
```

For data with `NAs` mean is `r round(mean_bydt,0)` and median is `r med_bydt`.  
For the imputed data mean is `r as.integer(mean_imp)` and median is `r as.integer(med_imp)`.

> Clearly both mean and median increased upon imputing the NAs in activity data

## Are there differences in activity patterns between weekdays and weekends?

1. Adding a new factor variable :
```{r weekdays}
weekdaynames <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
imputed <- imputed %>% mutate(weekend = weekdays(date) %in% weekdaynames)
```

2. Panel plot of daily average steps as function of intervals in weekends and weekdays:
```{r plot-final,message=F,fig.align='center',fig.height=8}
by_inter_imp <- imputed %>% 
           group_by(interval,weekend) %>%
           summarise(tot_st = mean(steps,na.rm=T))
by_inter_imp <- transform(by_inter_imp,weekend = factor(weekend,labels = c("weekend","weekday")))
xyplot(tot_st ~ interval | weekend ,data = by_inter_imp,layout = c(1,2),
       xlab = "interval",ylab = "steps",type = "l")
```